---
title: "Amphibians Final"
author: "Hannah Hagearty"
format: html
editor: visual
---

## Frogs, Salamanders and Toads YAY!!!

#load in the packages needed

```{r}
library(tidyverse)
library(ggplot2)
library(knitr)
library(dplyr)
```

#loading in data

```{r}
amph_data <- read.csv("Amphibians_on_NPS_Lands_20231201.csv")
view(amph_data)

```

#clean up data

```{r}
str(amph_data)



```

#Make a map with individuals by points - Common Name (number of ones that occur in a state), graphed on the states the parks occur in, showing how many species of amphibian were recorded per park in the present day. Will switch up by by Park occurance status.

```{r}
# I got help from Chat GTP on this question, mostly i would get an error and ask it how to fix it, and I also asked it how to get a graph template, hope that is ok?

library(ggplot2)
library(maps)
library(dplyr)

# Get the map of the USA - with help from chat GTP
usa_map <- map_data("state")

View(usa_map)

map_data <- amph_data %>%
  select(State, Common_Names, Park_OccurrenceStatus) %>%
  filter(Park_OccurrenceStatus == 'Present') %>% #filtering for present information
  group_by(State) %>%
  summarise(count = n())
# ok so this is taking the number of common names found in each state and it is going to hopefully graph them on a map of the united states, to show species diversity of amphibians in national parks by state.
 
state_abbreviation_to_name <- data.frame(
  abbreviation = c("AL", "AK", "AZ", "AR", "CA", "CO", "CT", "DE", "FL", "GA", 
                   "HI", "ID", "IL", "IN", "IA", "KS", "KY", "LA", "ME", "MD", 
                   "MA", "MI", "MN", "MS", "MO", "MT", "NE", "NV", "NH", "NJ", 
                   "NM", "NY", "NC", "ND", "OH", "OK", "OR", "PA", "RI", "SC", 
                   "SD", "TN", "TX", "UT", "VT", "VA", "WA", "WV", "WI", "WY"),
  name = c("alabama", "alaska", "arizona", "arkansas", "california", "colorado", 
           "connecticut", "delaware", "florida", "georgia", "hawaii", "idaho", 
           "illinois", "indiana", "iowa", "kansas", "kentucky", "louisiana", 
           "maine", "maryland", "massachusetts", "michigan", "minnesota", 
           "mississippi", "missouri", "montana", "nebraska", "nevada", "new hampshire", 
           "new jersey", "new mexico", "new york", "north carolina", "north dakota", 
           "ohio", "oklahoma", "oregon", "pennsylvania", "rhode island", "south carolina", 
           "south dakota", "tennessee", "texas", "utah", "vermont", "virginia", 
           "washington", "west virginia", "wisconsin", "wyoming")
) # change the labels in the maps_data, to lowercase state names because i need that to merge with the stupid map :(

map_data <- map_data %>%
  left_join(state_abbreviation_to_name, by = c("State" = "abbreviation"))

View(map_data)

map_data <- map_data %>%
  drop_na() #get rid of stuff that is not labeled as a state (like territories and islands)
  

#couldn't get this to work, so asked chat GTP and it said that my merge was not working because the states were named different abbreviations, so this should change the map data to how the map labels things.

View(map_data)
 
usa_map_data <- usa_map %>%
  left_join(map_data, by = c("region" = "name")) #join them together

View(usa_map_data)


#get the map into R.
#got help from chat gtp on this
#ggplot(data = usa_map, aes(x = long, y = lat, group = group)) +
#  geom_polygon(fill = "lightgray", color = "black") + 
  
ggplot(data = usa_map_data, aes(x = long, y = lat, group = group, fill = count)) +
  geom_polygon(color = "black") +  # Borders of the states
  scale_fill_viridis_c(option = "D") +  # said this one was color blind friendly
                  
  #scale_fill_continuous() +
  labs(title = "Species Diversity of Amphibians in National Parks by State - Present day", fill = "Count") +
  theme_minimal() +
  theme(legend.position = "right")+
  ylab("Latitude") +
  xlab("Longitude")


```
```{r}
# got help from notes and chat GTP on how to run this statistical test and on which statistical test to pick, because initially i thought I would run a T-test.

library(tidyverse)
library(gt)
library(DAAG)
library(car)
library(multcomp)

# because I have a categorical vaiable and the region column had more then two options, I need to run an anova test instead of just a t test.
anova_result <- aov(count ~ region, data = usa_map_data)
summary(anova_result)

#post hock test to see if there is significant differences in the data
tukey_result <- TukeyHSD(anova_result)
tukey_result




```

```{r}
usa_map_h <- map_data("state")

View(usa_map_h)

map_data_h <- amph_data %>%
  select(State, Common_Names, Park_OccurrenceStatus) %>%
  filter(Park_OccurrenceStatus == 'Historical') %>% #filtering for present day information
  group_by(State) %>%
  summarise(count = n())
# ok so this is taking the number of common names found in each state and it is going to hopefully graph them on a map of the united states, to show species diversity of amphibians in national parks by state.
 
state_abbreviation_to_name_h <- data.frame(
  abbreviation = c("AL", "AK", "AZ", "AR", "CA", "CO", "CT", "DE", "FL", "GA", 
                   "HI", "ID", "IL", "IN", "IA", "KS", "KY", "LA", "ME", "MD", 
                   "MA", "MI", "MN", "MS", "MO", "MT", "NE", "NV", "NH", "NJ", 
                   "NM", "NY", "NC", "ND", "OH", "OK", "OR", "PA", "RI", "SC", 
                   "SD", "TN", "TX", "UT", "VT", "VA", "WA", "WV", "WI", "WY"),
  name = c("alabama", "alaska", "arizona", "arkansas", "california", "colorado", 
           "connecticut", "delaware", "florida", "georgia", "hawaii", "idaho", 
           "illinois", "indiana", "iowa", "kansas", "kentucky", "louisiana", 
           "maine", "maryland", "massachusetts", "michigan", "minnesota", 
           "mississippi", "missouri", "montana", "nebraska", "nevada", "new hampshire", 
           "new jersey", "new mexico", "new york", "north carolina", "north dakota", 
           "ohio", "oklahoma", "oregon", "pennsylvania", "rhode island", "south carolina", 
           "south dakota", "tennessee", "texas", "utah", "vermont", "virginia", 
           "washington", "west virginia", "wisconsin", "wyoming")
) # change the labels in the maps_data, to lowercase state names because i need that to merge with the stupid map :(

map_data_h <- map_data_h %>%
  left_join(state_abbreviation_to_name_h, by = c("State" = "abbreviation"))

View(map_data_h)

map_data_h <- map_data_h %>%
  drop_na() #get rid of stuff that is not labeled as a state (like territories and islands)
  

#couldn't get this to work, so asked chat GTP and it said that my merge was not working because the states were named different abbreviations, so this should change the map data to how the map labels things.

View(map_data_h)
 
usa_map_data_h <- usa_map_h %>%
  left_join(map_data_h, by = c("region" = "name")) #join them together

View(usa_map_data_h)


#get the map into R.
#got help from chat gtp on this
#ggplot(data = usa_map, aes(x = long, y = lat, group = group)) +
#  geom_polygon(fill = "lightgray", color = "black") + 
  
ggplot(data = usa_map_data_h, aes(x = long, y = lat, group = group, fill = count)) +
  geom_polygon(color = "black") +  # Borders of the states
  scale_fill_viridis_c(option = "D") +  # said this one was color blind friendly
                  
  #scale_fill_continuous() +
  labs(title = "Species Diversity of Amphibians in National Parks by State - Historical", fill = "Count") +
  theme_minimal() +
  theme(legend.position = "right")+
  ylab("Latitude") +
  xlab("Longitude")



```

```{r}
usa_map_Adj <- map_data("state")

View(usa_map_Adj)

map_data_Adj <- amph_data %>%
  select(State, Common_Names, Park_OccurrenceStatus) %>%
  filter(Park_OccurrenceStatus == 'Adjacent') %>% #filtering for present day information
  group_by(State) %>%
  summarise(count = n())
# ok so this is taking the number of common names found in each state and it is going to hopefully graph them on a map of the united states, to show species diversity of amphibians in national parks by state.
 
state_abbreviation_to_name_Adj <- data.frame(
  abbreviation = c("AL", "AK", "AZ", "AR", "CA", "CO", "CT", "DE", "FL", "GA", 
                   "HI", "ID", "IL", "IN", "IA", "KS", "KY", "LA", "ME", "MD", 
                   "MA", "MI", "MN", "MS", "MO", "MT", "NE", "NV", "NH", "NJ", 
                   "NM", "NY", "NC", "ND", "OH", "OK", "OR", "PA", "RI", "SC", 
                   "SD", "TN", "TX", "UT", "VT", "VA", "WA", "WV", "WI", "WY"),
  name = c("alabama", "alaska", "arizona", "arkansas", "california", "colorado", 
           "connecticut", "delaware", "florida", "georgia", "hawaii", "idaho", 
           "illinois", "indiana", "iowa", "kansas", "kentucky", "louisiana", 
           "maine", "maryland", "massachusetts", "michigan", "minnesota", 
           "mississippi", "missouri", "montana", "nebraska", "nevada", "new hampshire", 
           "new jersey", "new mexico", "new york", "north carolina", "north dakota", 
           "ohio", "oklahoma", "oregon", "pennsylvania", "rhode island", "south carolina", 
           "south dakota", "tennessee", "texas", "utah", "vermont", "virginia", 
           "washington", "west virginia", "wisconsin", "wyoming")
) # change the labels in the maps_data, to lowercase state names because i need that to merge with the stupid map :(

map_data_Adj <- map_data_Adj %>%
  left_join(state_abbreviation_to_name_Adj, by = c("State" = "abbreviation"))

View(map_data_Adj)

map_data_Adj <- map_data_Adj %>%
  drop_na() #get rid of stuff that is not labeled as a state (like territories and islands)
  

#couldn't get this to work, so asked chat GTP and it said that my merge was not working because the states were named different abbreviations, so this should change the map data to how the map labels things.

View(map_data_Adj)
 
usa_map_data_Adj <- usa_map_Adj %>%
  left_join(map_data_Adj, by = c("region" = "name")) #join them together

View(usa_map_data_Adj)


#get the map into R.
#got help from chat gtp on this
#ggplot(data = usa_map, aes(x = long, y = lat, group = group)) +
#  geom_polygon(fill = "lightgray", color = "black") + 
  
ggplot(data = usa_map_data_Adj, aes(x = long, y = lat, group = group, fill = count)) +
  geom_polygon(color = "black") +  # Borders of the states
  scale_fill_viridis_c(option = "D") +  # said this one was color blind friendly
                  
  #scale_fill_continuous() +
  labs(title = "Species Diversity of Amphibians in National Parks by State - Adjacent", fill = "Count") +
  theme_minimal() +
  theme(legend.position = "right")+
  ylab("Latitude") +
  xlab("Longitude")






```

```{r}
usa_map_PO <- map_data("state")

View(usa_map_PO)

map_data_PO <- amph_data %>%
  select(State, Common_Names, Park_OccurrenceStatus) %>%
  filter(Park_OccurrenceStatus == 'Possible') %>% #filtering for present day information
  group_by(State) %>%
  summarise(count = n())
# ok so this is taking the number of common names found in each state and it is going to hopefully graph them on a map of the united states, to show species diversity of amphibians in national parks by state.
 
state_abbreviation_to_name_Adj <- data.frame(
  abbreviation = c("AL", "AK", "AZ", "AR", "CA", "CO", "CT", "DE", "FL", "GA", 
                   "HI", "ID", "IL", "IN", "IA", "KS", "KY", "LA", "ME", "MD", 
                   "MA", "MI", "MN", "MS", "MO", "MT", "NE", "NV", "NH", "NJ", 
                   "NM", "NY", "NC", "ND", "OH", "OK", "OR", "PA", "RI", "SC", 
                   "SD", "TN", "TX", "UT", "VT", "VA", "WA", "WV", "WI", "WY"),
  name = c("alabama", "alaska", "arizona", "arkansas", "california", "colorado", 
           "connecticut", "delaware", "florida", "georgia", "hawaii", "idaho", 
           "illinois", "indiana", "iowa", "kansas", "kentucky", "louisiana", 
           "maine", "maryland", "massachusetts", "michigan", "minnesota", 
           "mississippi", "missouri", "montana", "nebraska", "nevada", "new hampshire", 
           "new jersey", "new mexico", "new york", "north carolina", "north dakota", 
           "ohio", "oklahoma", "oregon", "pennsylvania", "rhode island", "south carolina", 
           "south dakota", "tennessee", "texas", "utah", "vermont", "virginia", 
           "washington", "west virginia", "wisconsin", "wyoming")
) # change the labels in the maps_data, to lowercase state names because i need that to merge with the stupid map :(

map_data_PO <- map_data_PO %>%
  left_join(state_abbreviation_to_name_Adj, by = c("State" = "abbreviation"))

View(map_data_PO)

map_data_PO <- map_data_PO %>%
  drop_na() #get rid of stuff that is not labeled as a state (like territories and islands)
  

#couldn't get this to work, so asked chat GTP and it said that my merge was not working because the states were named different abbreviations, so this should change the map data to how the map labels things.

View(map_data_PO)
 
usa_map_data_PO <- usa_map_PO %>%
  left_join(map_data_PO, by = c("region" = "name")) #join them together

View(usa_map_data_PO)


#get the map into R.
#got help from chat gtp on this
#ggplot(data = usa_map, aes(x = long, y = lat, group = group)) +
#  geom_polygon(fill = "lightgray", color = "black") + 
  
ggplot(data = usa_map_data_PO, aes(x = long, y = lat, group = group, fill = count)) +
  geom_polygon(color = "black") +  # Borders of the states
  scale_fill_viridis_c(option = "D") +  # said this one was color blind friendly
                  
  #scale_fill_continuous() +
  labs(title = "Species Diversity of Amphibians in National Parks by State - Possible", fill = "Count") +
  theme_minimal() +
  theme(legend.position = "right")+
  ylab("Latitude") +
  xlab("Longitude")



```

# Globabl Conservation Status By Family

```{r}
#Trying to graph Global Conservation Status by Family of amphibian, color indicates native or not

Global_conservation_status_rank_by_family <- amph_data %>%
  select(GRank, Family, Nativeness) 


view(Global_conservation_status_rank_by_family)

ggplot(data = Global_conservation_status_rank_by_family, aes(x = GRank , y = Family, color = Nativeness )) +
  geom_point()
```

#These are the Stacked Bar Plots that each go with one of the Maps.

```{r}
#Family diversity by State, this should go with the Map that is Adjacent

map_data_Adj_bar <- amph_data %>%
  select(State, Common_Names, Park_OccurrenceStatus, Family) %>%
  filter(Park_OccurrenceStatus == 'Adjacent') %>% #filtering for Adjacent information
  filter(State != "Mariana Islands") %>%
  filter( State != "Trail (ME to GA)") %>% # got rid of stuff on a trail or stuff on a random island
  group_by(State, Family) %>%
  summarise(count = n() )

view(map_data_Adj_bar)


ggplot(data = map_data_Adj_bar, aes(fill= Family , y=count, x=State)) + 
    geom_bar(position="stack", stat="identity") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

```{r}
#Family diversity by State, this should go with the Map that is Present

map_data_bar <- amph_data %>%
  select(State, Common_Names, Park_OccurrenceStatus, Family) %>%
  filter(Park_OccurrenceStatus == 'Present') %>%#filtering for present information
  filter(State != "Mariana Islands") %>%
  filter( State != "Trail (ME to GA)") %>% # got rid of stuff on a trail or stuff on a random island
  group_by(State, Family) %>%
  summarise(count = n() )

view(map_data_bar)


ggplot(data = map_data_bar, aes(fill= Family , y=count, x=State)) + 
    geom_bar(position="stack", stat="identity") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))


```

```{r}
map_data_his_bar <- amph_data %>%
  select(State, Common_Names, Park_OccurrenceStatus, Family) %>%
  filter(Park_OccurrenceStatus == 'Historical') %>%#filtering for Historical information
  filter(State != "Mariana Islands") %>%
  filter( State != "Trail (ME to GA)") %>% # got rid of stuff on a trail or stuff on a random island
  group_by(State, Family) %>%
  summarise(count = n() )

view(map_data_his_bar)


ggplot(data = map_data_his_bar, aes(fill= Family , y=count, x=State)) + 
    geom_bar(position="stack", stat="identity") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

```{r}
map_data_pos_bar <- amph_data %>%
  select(State, Common_Names, Park_OccurrenceStatus, Family) %>%
  filter(Park_OccurrenceStatus == 'Possible') %>% #filtering for Possible information
  filter(State != "Mariana Islands") %>%
  filter( State != "Trail (ME to GA)") %>% # got rid of stuff on a trail or stuff on a random island
  group_by(State, Family) %>%
  summarise(count = n() )

view(map_data_pos_bar)


ggplot(data = map_data_pos_bar, aes(fill= Family , y=count, x=State)) + 
    geom_bar(position="stack", stat="identity") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))


```

```{r}
#working on this

#library(dplyr)
#library(lubridate)


#class(amph_data$Verified_Date)
#amph_data$Verified_Date <- as.Date(amph_data$Verified_Date) # chat gtp helped me figure out that my dates were being stored as characters and thats why the below code was not working.
#amph_data$Verified_Date <- mdy(amph_data$Verified_Date)
 #view(amph_data) # had to change up the order of day month and year so that everything got pulled to the correct columns.

#native <- amph_data %>%
#  select(State, Nativeness, Verified_Date) %>%
#  mutate(
 #   YEAR = year(Verified_Date),
 #   MONTH = month(Verified_Date),
 #   DAY = day(Verified_Date)
 # ) %>%
 # group_by(State, Nativeness) %>%
 # summarise(count = n())
 

#view(native)


#ggplot(data = native, aes(y=count, x=State, color = Nativeness)) + 
#    geom_point() +
#  theme(axis.text.x = element_text(angle = 90, hjust = 1))


#native <- amph_data %>%
#  select(State, Nativeness, Verified_Date) %>%
#  mutate(
#    YEAR = year(Verified_Date),
#    MONTH = month(Verified_Date),
 #   DAY = day(Verified_Date)
 # ) %>%
 # group_by(State, Nativeness, Year) %>%
 # summarise(count = n())
 



```

```{r}

library(leaflet)


# this was to get the names of the parks
park_names <- amph_data %>%
  select(Park_Name) %>%
  unique()

view(park_names)

write.csv(park_names, "parknames.csv", row.names = FALSE)

library(tidygeocoder)
library(dplyr)
library(tidyr)

# Load your CSV file
file_path <- "parknames.csv"  # Replace with your file path
data <- read.csv(file_path)

# Geocode the locations
geocoded_data <- data %>%
  geocode(Park_Name, method = "osm", lat = Latitude, long = Longitude)

# Save the output to a new CSV file
output_path <- "parknames_with_coordinates.csv"
write.csv(geocoded_data, output_path, row.names = FALSE)

print(paste("Geocoding completed. Results saved to", output_path))


```

```{r}
library(tidyverse)
library(maps)

world_maps <- map_data("world")

view(world_maps)

USA <- world_maps %>%
  filter(region == "USA")

library(readr)
parknames_with_coordinates <- read_csv("parknames_with_coordinates.csv")

nationalparks <- data.frame(parknames_with_coordinates)
view(nationalparks)


ggplot(data = USA, aes(x = long, y = lat, group = group)) +
  geom_polygon(fill = "gray", color = "white") +  # Draw the map
  coord_quickmap() +  # Ensure correct aspect ratio
  geom_point(data = nationalparks, aes(x = Longitude, y = Latitude, group = 2), color = "red")

```

```{r}
# This figure shows the locations of all of the national parks that had amphibians recorded in them.

library(leaflet)

leaflet(nationalparks) %>%
  addTiles() %>%  # Add a base map
  addMarkers(
    lng = ~Longitude,  # Use the Longitude column for marker positions
    lat = ~Latitude,   # Use the Latitude column for marker positions
    popup = ~ Park_Name)     # Optional: Use a column named 'popup' for labels


```

```{r}
#This graph the native vs. not native species count found in each state. Park occurrence is present.

nativevsnot_data_bar <- amph_data %>%
  select(State, Common_Names, Park_OccurrenceStatus, Family, Nativeness) %>%
  filter(Park_OccurrenceStatus == 'Present') %>%#filtering for present information
  filter(State != "Mariana Islands") %>%
  filter( State != "Trail (ME to GA)") %>% # got rid of stuff on a trail or stuff on a random island
  group_by(State, Nativeness) %>%
  summarise(count = n() )

view(nativevsnot_data_bar)


ggplot(data = nativevsnot_data_bar, aes(fill= Nativeness , y=count, x=State)) + 
    geom_bar(position="stack", stat="identity") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))



```





